---
title: "00 - Setup - Installing Required Software"
author: "Matthias Grenié & Marten Winter"
date: "Monday April 29th 2022"
output:
  html_document:
    toc: yes
bibliography: bibliography.bib
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```


# Setup

## RStudio

Double-click on the `.Rproj` file which is as the root of the folder you downloaded. This should open a new session in RStudio that will start at the root of the project folder.

## Following the tutorial

To follow the tutorial you have plenty of options:

1. It is accessible online at https://rekyt.github.io/biodiversity_facets_tutorial/ click on `Tutorial` in the navigation bar to access the tutorial.
1. You can access open the `diversity_facets_tutorial.Rmd` in RStudio to follow along.
1. You can also open the `diversity_facets_tutorial.R` in RStudio to get only the code to execute exactly the codes here.

## Installing needed packages

We will use some specific packages in the rest of the tutorial. To make sure you have them please run the following command:

```{r install-pkgs, eval = FALSE}
install.packages(
  c("ade4",
    "ape",
    "FD",
    "fundiversity",
    "ggplot2",
    "ggspatial",
    "performance",
    "picante",
    "rnaturalearth",
    "sf")
)
```

> **Note**: If you encounter trouble while installing software notify me at the beginning of the practical session and I'll come to you to solve the issue
