---
title: "Answers to questions asked in the tutorial"
author: "Matthias Grenié & Marten Winter"
date: "Monday July 5th 2021"
output:
  html_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

Here are answers regarding all the questions asked in the tutorial.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Loading the data (**Q1**-**Q6**)

## Summarizing the data

* **Q1**: How many plots were sampled?

```{r q1}
nrow(plot_data)
```

**Answer**: There are 180 sampled plots. 


* **Q2**: How many species are there in the dataset?

```{r q2}
nrow(species_traits)
```

**Answer**: There are 691 species in the dataset.


* **Q3**: How many traits are available?

```{r q3}
head(species_traits, 2)
```

**Answer**: There are 13 traits in the dataset.


* **Q4**: How many of them are continuous? How many of them are discrete?

**Answer**: Looking at the trait table there 3 continuous traits (`height`, `sla`, and `wood.dens`) and 10 that are discrete. (Referring to the README file we see that the `seed` column contains indeed a categorical trait). 

* **Q5**: What is the most numerous family among all observed species?

```{r q5}
summary(species_traits)
```

**Answer**: The `summary()` function displays the number of observations in each categories ordered by the most abundant category. The `family` column shows that the "Leguminosae" (a.k.a, Fabaceae) is the most numerous in the dataset.

* **Q6**: What is the most numerous genus?

**Answer**: Similarly the most numerous genus is "Unknown", which corresponds to individual that weren't identified at the genus level.

# Functional diversity (**Q7**-**Q19**)

## Computing Biomass-weighted mean traits per plot (**Q7**-**Q10**)

* **Q7**: How would you describe the relationship between the different CWMs and forest loss?

**Answer**: We're eye-balling the relationships but it seems that:

* The CWM-Height seems to decrease with forest loss.
* The CWM-SLA seems to increase with forest loss.
* There seem to be no relationship between the CWM-wood density and forest loss. 

* **Q8**: Can you test the correlation using the function `cor.test()` and does it support your previous statements?

```{r q8}
cor.test(cwm_env$forestloss17, cwm_env$height)
cor.test(cwm_env$forestloss17, cwm_env$sla)
cor.test(cwm_env$forestloss17, cwm_env$wood.dens)
```

**Answer**:

* We observe a negative correlation (-0.48) between CWM-Height and forest loss.
* We observe a positive correlation (0.48) between CWM-SLA and forest loss.
* We observe a slight negative correlation (-0.24) between CWM-wood density and forest loss. 

* **Q9**: How would you describe the understorey vegetation changes with increasing forest loss?

**Answer**: The CWM values reflect the dominant trait of the community. This means that with forest loss we observe smaller species, species with higher SLA, and species with lower wood density. Linking back to ecology, this is probably due to a change in understorey vegetation type from woody species in unlogged area to more herbaceous species in heavily logged area.

* **Q10**: How does this observation compare to above description of the change of understorey vegetation along the forest loss gradient?

```{r q10}
cor.test(non_quanti_cwm$forestloss17, non_quanti_cwm$woody_no)
```

**Answer**: The `woody_no` CWM represents the proportion of non-woody species in the community. We observe a positive medium correlation between this CWM and forest loss which tells us that non-woody species are more frequent in understorey vegetation with increasing forest loss.

## Building the functional space (**Q11**-**Q12**)

* **Q11**: Using the metadata available in the `README.txt` file, what is the meaning of the `pgf` column?

**Answer**: The `README.txt` file (in the `data/doi_10.5061_dryad.f77p7__v1/` folders) describes the full dataset with the meaning of all columns. It says:

> pgf: Plant growth form: A = fern, B = graminoid, C = forb, D = herbaceous climber, E = herbaceous shrub, F = tree sapling, G = woody climber, H = woody shrub, na = indeterminate

So the `pgf` column describes the plant growth form.

* **Q12**: How do you interpret the PCoA results given your answer to the previous question?

**Answer**: The PCoA shows groups of species according to their plant growth form. Especially the category "F" (tree saplings) seem to differentiate from all other growth forms. This means that tree saplings have distinct traits from all other growth form groups.


## Computing functional diversity indices (**Q13**-**Q15**)

* **Q13**: How would you describe the relationships between functional diversity and forest loss and road density?
* **Q14**: Using the plot generated by the code beneath how could you describe the relationships between the three different functional diversity indices we computed?
* **Q15**: How does the relationship between indices with species richness compare with the one observed with total biomass values? (You can use the function `cor.test()` if you want to test the association)

## Null modeling (**Q16**-**Q19**)

* **Q16**: How would describe verbally the position of the observed value of FRic for site "a100f177r" compared to the null distribution?
* **Q17**: What the quantile of the observed FRic value in the end?
* **Q18**: Using the `subset()` function with the greater (or equal) than `>=` and the lower (or equal) than `<=`, can you determine how many sites show a significant deviation from the null observation? (absolute SES >= 2)
* **Q19**: Using similar code as used for observed values, what are the relationships between SES values and forest loss?

# Phylogenetic diversity (**Q20**-**Q27**)

## Getting the phylogenetic tree (**Q20**-**Q21**)

* **Q20**: How many taxa are in the phylogenetic tree?
* **Q21**: How does this number compare to the number of taxa found in the dataset?

## Computing phylogenetic diversity indices (**Q22**-**Q24**)

* **Q22**: What do you notice with the species names? Especially compared to the ones available in `species_traits`.
* **Q23**: What are the relationships between weighted and unweighted version of the MPD?
* **Q24**: What are the relationships between MPD and taxa richness? And with forest loss? Plot these relationships to visualize them and use the `cor.test()` function to validate your observations.

## Null modeling (**Q25**-**Q27**)

* **Q25**: Explain what does the column `mpd.obs.z` means? How does this compare with the SES values we computed for functional diversity indices?
* **Q26**: How does the standardized value relates with taxa richness?
* **Q27**: What are the relationships between MPD values considering null models and forest loss? Visualize the relationships with the `plot()` function, validate your observations with the `cor.test()` function.

# Comparing facets (**Q28**)

* **Q28**: How are related are observed values of functional diversity and phylogenetic diversity? What about the SESs?

# Modelling the effect of logging (**Q29**-**Q31**)

## Single-predictor models (**Q29**-**Q30**)

* **Q29**: How would you qualify the effect of forest loss on the taxa richness?
* **Q30**: With the same formula build similar models with the other predictors `roaddensprim` and `roaddistprim`. How do they compare with forest loss?

## Multi-predictors models (**Q31**)

* **Q31**: What can you say about the effect of the disturbances on the different diversity metrics? What are the explanatory power of our models?
